---
title: "Lab3b_SNC"
author: "Stephanie Call"
date: "9/13/2020"
output: html_document
bibliography: data/references.bib
---

# This tutorial is from the Data Carpentry "Manipulating, analyzing and exporting data with tidyverse" tutorial (@DataCarpentryT2)

## Import the tidyverse package and read the data to memory
```{r library}
options(max.print = 500) # Make all data outputs no more than 500 observations
library("tidyverse")

# Read the surveys spreadsheet using readr's function read_csv()
surveys <- read_csv("data/portal_data_joined.csv")
# Now look at the data using str() and preview using View()
str(surveys)
# Similar to before, except more details, including length of each
# variable vector and list of attributes that specified each variable's
# (column's) data type in readr. Also, instead of data frame, is now 
# a tibble object. It's also much more colorful and easier to read
View(surveys)
# This looks the same as before - opens spreadsheet in another window
```

## Selecting columns and filtering rows 

```{r selectFilter}
# Select columns (variables) using select() function from dplyr
select(surveys, plot_id, species_id, weight)
# To select all columns except the ones listed, use -
select(surveys, -record_id, -species_id)
# To select rows (observations) based on criterion (or criteria),
# use the filter() function from dplyr
filter(surveys, year == 1995)
```

## Pipes

```{r Pipes}
# To select and filter at the same time, there are some options
# Can create a temporary intermediate data frame -
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)
head(surveys_sml)
# However, this is a pain to keep track of and clutters the environment.
# Alternatively, can nest functions in each other
surveys_sml <- select(filter(surveys, weight < 5), species_id, sex, weight)
head(surveys_sml)
# Can be useful for small and simple analysis, but is difficult to 
# read if too many are nested inside of each other. 
# Finally, can use pipes, which take the output of one function and inputs
# it directly to the next using the pipe operator (%>%) from magrittr
surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)
# Objects on the left are passed to the first argument of the 
# function to the right.
# Can simply assign the new object as usual with the assignment operator
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
surveys_sml
```

## Challenge Question 1

```{r Q1}
# Use pipes to subset surveys for animal data before 1995 with just 
# the year, sex, and weight 
surveys_pre1995 <- surveys %>% 
  filter(year < 1995) %>% 
  select(year, sex, weight)
surveys_pre1995
```

## Mutate

```{r mutate}
# Can use the dplyr function mutate() to add additional columns based
# on values in pre-existing columns
# Add a new weight column in kg -
surveys %>% 
  mutate(weight_kg = weight/1000)
# Can add multiple columns at once, even using new columns made
# in the same call
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)
# Pipes also work with non-dplyr functions - 
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
# Filter the data to remove unknown weights before adding the new column -
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

## Challenge Question 2

```{r Q2}
# 1. Make a new data frame (technically tribble) from surveys with
# only the species ID and hindfoot variables, with the hindfoot in
# cm (originally in mm), lengths less than 3, and no NAs.
surveys_hindfoot_sml <- surveys %>% 
  mutate(hindfoot_cm = hindfoot_length/10) %>% 
  filter(!is.na(hindfoot_cm) & hindfoot_cm < 3) %>% 
  select(species_id, hindfoot_cm)
head(surveys_hindfoot_sml)
```

## Split-apply-combine data analysis and the summarize() function

```{r group_and_summarize}
# group_by() from dplyr can be used to group categorical variables 
# that can then be summarized with the summarize() function to 
# display the statistics for each group in a sinlge line -
surveys %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE))
# Can also group by multiple columns (variables) 
# by listing multiple in group_by(). Note that it groups by 
# all possible variable combinations
surveys %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>% 
  tail()
# See that there are many NA sexes and NaN weights, meaning the 
# weights weren't taken before the animal escaped. 
# To remove these data, use the filter() function before analysis
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
# Can display as much data as you want from output using print() function -
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>% 
  print(n = 15)
# Can summarize multiple variables at the same time with multiple statistics 
# by listing multiple statistics (calculations) in summarize()
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
# Can use the arrange() function to sort the output for easier inspection
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
# Will sort in ascending order by default. For descending, 
# must use the desc() function
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))

```

## Counting

```{r count}
# To count the number of observations (rows) for each factor/combination
# of factors, use the dplyr count() function
count(surveys, sex)
# This is essentially a summary of the group_by() function, equivalent to
surveys %>% 
  group_by(sex) %>% 
  summarise(count = n())
# Can sort output from count using the optional sort output
count(surveys, sex, sort = TRUE) # Sorts in descending order by default
# To specify multiple factors at one, specify as separate arguments
count(surveys, sex, species)
# Can then sort these data using arrange(). Can also sort by multiple
# criteria in arrange() function by listing multiple criteria 
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))
```

## Challenge Question 3

```{r Q3}

# 1. How many animals were caught in each plot_type surveyed?
count(surveys, plot_type)

# 2. Use group_by() and summarize() to find the mean, min, and max
# hindfoot length for each species. Add total number of observations
surveys %>% 
  group_by(species_id) %>% 
  filter(!is.na(hindfoot_length)) %>% 
  summarize(mean_hindfoot_length = mean(hindfoot_length),
            minimum_hindfoot_length = min(hindfoot_length),
            maximum_hindfoot_length = max(hindfoot_length),
            count = n())

# 3. What was the heaviest animal measured each year? Return 
# the columns year, genus, species_id, and weight
surveys %>% 
  group_by(year) %>%
  filter(!is.na(weight)) %>% 
  filter(weight == max(weight)) %>% 
  select(year, genus, species_id, weight) %>% 
  arrange(year)
```

## Reshaping with gather() and spread()

```{r reshape}
# Spreading 

# Spreading takes a column variable and spreads it into different 
# columns (each entry is now a separate variable) with each entry 
# being a different value from the input data.
# spread() takes 3 main arguments - data, key, and value
# The observations stay the same while the data are reorganized
# based on the input key vector and the corresponding values
# Ex - Transform surveys to extract the mean weight of each genus 
# in each plot
surveys_gw <- surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(plot_id, genus) %>% 
  summarize(mean_weight = mean(weight))
str(surveys_gw)   #Inspect the extracted data
# Now, use spread to reorganize the data by genus (key) for the 
# mean weights (value)
surveys_spread <- surveys_gw %>% 
  spread(key = genus, value = mean_weight)
str(surveys_spread)
head(surveys_spread)
# Now can make plots for comparisons, after filling in the missing
# values with the fill argument 
surveys_gw %>% 
  spread(genus, mean_weight, fill = 0) %>% 
  head()


# Gathering

# This is essentially the opposite of spread(), where the column
# names (variables) are gathered together into a single column
# (variable for the overall data) with each name now a factor.
# The value for each entry is then put into another new column
# gather() takes four main arguments - data, key, values, and 
# the names of the columns to fill in the key variable.
# The final argument can also be used to exclude values for key
# Ex - Recreate surveys_gw from surveys_spread - 
surveys_gather <- surveys_spread %>% 
  gather(key = 'genus', value = 'mean_weight', -plot_id)
str(surveys_gather)
str(surveys_gw)
# Note that surveys_gather has NA for all genera in every plot, 
# so it has more observations than surveys_gw. 
# This balances out the dataset so every combination has an entry.
# Spreading and gathering is a good way to balance a dataset.
# gather() can also be used to specify which columns to use in
# the fourth place argument. The colon operator can be easily used
# to select adjacent column values. For example, 
surveys_spread %>% 
  gather(key = "genus", value = "mean_weight", Baiomys:Spermophilus) %>%
  head()
```

## Challenge Question 4

```{r Q4}
# 1. Spread surveys data with year as columns, plot_id as rows, and 
#    the number of genera per plot as values
surveys_pg <- surveys %>% 
  filter(!is.na(year)) %>% # filter out any entries without a year, just in case
  group_by(plot_id, year) %>% 
  summarize(num_genera = n_distinct(genus))
head(surveys_pg) # Check that the data were organized properly
# Now reshape using spread() 
surveys_pg_spread <- surveys_pg %>% 
  spread(key = "year", value = "num_genera", fill = 0)
head(surveys_pg_spread)

# 2. Gather the data again with the gather function so each row 
# is a unique plot_id by year combination
surveys_pg_gather <- surveys_pg_spread %>% 
  gather(key = 'year', value = 'num_genera', -plot_id)
head(surveys_pg_gather)

# 3. Gather the measurement data from survey into a dataset that 
# specified the measurement type and value
surveys_measure <- surveys %>% 
  gather(key = 'measurement', value = 'value', hindfoot_length, weight)
head(surveys_measure)

# 4. Calculate the average of each measurement for each year for each
# plot_type, then spread the data into a data set by measurement type
# (hindfoot_length and weight).
surveys_measure_avg <- surveys_measure %>% 
  filter(!is.na(value)) %>%  # This is a very important step
  group_by(measurement, year, plot_type) %>% 
  summarize(average = mean(value))
head(surveys_measure_avg) # Check that the data were extracted properly
surveys_measure_avg_spread <- surveys_measure_avg %>% 
  spread(measurement, average)
surveys_measure_avg_spread
```

## Exporting data

```{r export}
# write_csv() writes data frames into CSV files
# Before doing so, will create a new data file to keep manipulated
# data sets separate from the raw data. Additionally, we will 
# clean up the data set to remove missing data.

# Remove the observations for animals that are missing weight, 
# hindfoot length, or sex data
surveys_complete <- surveys %>% 
  filter(!is.na(weight), !is.na(sex), !is.na(hindfoot_length))

# Now remove the species that are relatively rare (observed < 50 times)
# First, create a data set that counts the number of species, 
# then filter out the rare ones
species_count <- surveys_complete %>% 
  count(species_id) %>% 
  filter(n >= 50)
surveys_complete <- surveys_complete %>% 
  filter(species_id %in% species_count$species_id)
# Check that the data were extracted properly -
dim(surveys_complete) # 30463 observations and 13 variables
# Save the manipulated data set into a new CSV file in the data folder
```

```{r writeOutput, eval = FALSE}
write_csv(surveys_complete, path = 'data/surveys_complete.csv')
```

## REFERENCES